Nitric-oxide-feedback to ciliary photoreceptor cells orchestrates UV avoidance

### 2023-05-08 Manuscript/Manuscript.Rmd
MS: Discussion edits [tbc]

In the present study, we explored the role of NO in the UV avoidance behaviour of Platynereis. We demonstrated that there are interneurons expressing NOS in the cPRC circuit and that NO is required for UV avoidance in behavioural experiments using NOS mutants. We identified the axon-localised type NIT-GC1 and the cilia-localised type NIT-GC2 expressed in the cPRCs as targets of NO and showed from cultured cell experiments that these produce cGMP in response to NO. We have shown by calcium imaging that INNOSs are activated in response to the hyperpolarising response of the cPRCs, that the NO produced feeds back to the cPRC via NIT-GC1 and induces a depolarising response in the cPRC, and that the depolarising response of the cPRC is involved in activating downstream INRGW. Lastly, we confirmed that RGWamide induces the downward swimming behaviour of larvae.

@@

cPRCs expressing UV-absorbing c-opsin1 have high resting calcium levels and show hyperpolarising and subsequent depolarising responses to UV light (405 nm) [@Veraszto2018]. In the present study, we have shown that the cells that strongly synapse with cPRCs are interneurons that specifically express NOS (INNOS). The fluorescence intensity of DAF-FM, which specifically detected INNOS, showed a transient tonic-like increase dependent on UV stimulation of the cPRCs. The graph shows a decrease in intensity after UV light is switched off, but the response of DAF-FM to NO is irreversible and this decrease is due to breaching by the relatively strong excitation light. Attempts to measure changes in the intensity of the calcium indicator RGECO1a simultaneously with DAF-FM imaging did not lead to detection of changes in fluorescence intensity compared to GCaMP6s and could only visualise the cPRCs. In calcium imaging using GCaMP6s, INNOS identified by the positional relationship between the cPRC and INNOS and the labelling of RYa depolarised in clear synchrony with the hyperpolarising response of the cPRCs. This tonic response may involve dopaminergic nerves based on single-cell data **(Figure 6)** [@puopolo2001][@zhang2007][@Randel2014a]. In the mammalian retina, it is known that signals from bipolar photoreceptor cells are transmitted to downstream NOS-expressing amacrine cells (NOAC), where NO is produced [@jacoby2018]. These suggest that UV stimulation of cPRCs causes the production of NO via a depolarising response of INNOS.

Our results suggest that NIT-GC1, which is specifically expressed in cPRCs, functions as a target for NO produced by INNOS and contributes to the regulation of the cPRC circuit transmission pathway via a depolarising response by increasing cGMP. In the mammalian retina, NO is an essential signalling molecule and one of the most well-studied [@Cudeiro1999]. It has been reported that NO produced in one of the retinal cells modulates cGMP-mediated neurotransmission by activating presynaptic neural sGCs (general target of NO) through retrograde signalling [@Vielma2014][@Wei2012]. In Platynereis cPRC, NIT-GC1 was also identified as having the property of producing cGMP in a NO-dependent manner. Furthermore, it is known that CNGAα, one of the cyclic nucleotide-gated channels, is specifically highly expressed in cPRC [@Tosches2014], and in INNOS-cPRC signalling, NO is implicated in the neuromodulation of the cPRC by retrograde signalling. NO is a molecule with a very short half-life, and when functioning as a signalling molecule, NOS and sGC are known to localise very close together at synapses between neurons [@burette2002][@garthwaite2015a]. In our study, we found 12 sGCs in Platynereis, but none expressed in the cPRC. Localisation analysis of NOS and NIT-GC1 by antibody staining showed that the two proteins localised very close together in the anterior neurosecretory plexus region. Calcium imaging results using NOS mutant and NIT-GC1 knockdown larvae suggest that NIT-GC1 activation induces a depolarising response in the cPRC.

Another type of interneuron that strongly synapses with the cPRC is the INRGWa, which synapses directly onto the ciliomotor neuron and is therefore considered to be a very important neuron in the control of larval behaviour [@Veraszto2018][@williams2017]. The present results suggest that the depolarising response of the cPRC due to NO feedback induces INRGW activation. RGWamide is also a neuropeptide that induces larval downward swimming quite strongly (unpublished? need to show data?), and the delayed downward swimming of NOS mutant larvae in response to UV stimulation may be responsible for the failure of INRGW activation to occur. In addition, in many marine invertebrates, NO has been reported to be involved in many aspects of settlement behaviour and metamorphosis induction [@leise2001; @bishop2003; @bishop2007; @pechenik2007; @castellano2014; @ueda2013; @ueda2014; @ueda2016; @yang2018; @zhu2020; @song2021]. The downward swimming behaviour of Platynereis larvae is the first step in the onset of settlement behaviour and metamorphosis. These suggest that downward swimming due to NO activation of INRGWs may trigger the onset of settlement behaviour and metamorphosis [@Conzelmann2013][@Williams2015]. These suggest that NO feedback signalling may act as a trigger for downward swimming via activation of INRGWs, which induces the initiation of settlement behaviour and metamorphosis.

An outstanding question is how the NO\~NIT-GC response alters neuronal dynamics. NO is a single molecule and, due to its short half-life, cannot be transmitted over long distances and has a limited signalling repertoire. It has therefore been considered unsuitable for generating diversity in signalling compared to other signalling molecules such as neuropeptides and classical neurotransmitters [@Jékely2021]. However, our results in the present study indicate that NO-induced activation of NIT-GC1 is involved in the regulation of neuronal circuits (cPRC pathway). The NIT domain was originally found in bacteria and the existence of GCs with NIT domains was first reported by Moroz et al in 2020 [@Shu2003; @Moroz2020]. It has been proposed that NIT domains in bacteria regulate cellular functions in response to changes in extracellular and intracellular nitrate and/or nitrite concentrations [@camargo2007]. In fact, NO is a free radical molecule, with a half-life of 5 msec in vivo, and is readily converted to nitrate and nitrite [@santos2011; @möller2019; @garthwaite2015a]. And nitrate and nitrite have been reported to accumulate in cells and tissues with high NOS activity in placozoans and cnidarians [@Moroz2004; @Moroz2020]. It is possible that the NIT domains of NIT-GC1 identified in this study may also be used for nitrate/nitrite sensing, as in bacteria. Furthermore, if they show different sensitivities to NO, nitrite and nitrate, it is possible that the temporal variation in their respective half-lives could give rise to a repertoire of activation timings [@lundberg2011]. In addition, we have found at least 15 NIT-GCs in the Platynereis transcriptome data, of which NIT-GC2 is highly expressed in the cPRC, in addition to NIT-GC1. Interestingly, NIT-GC2 showed a very different localisation pattern to NIT-GC1, and furthermore, the results of calcium imaging of cPRC in response to UV stimulation using NIT-GC2 knockdown larvae were the different dynamics (such that hyperpolarisation was suppressed) with NIT-GC1 knockdown. It was not clear from the present analysis whether this difference was due to differences in the localisation or the property of the NIT-GCs. However, the diversification and spatially different localisation patterns of NIT-GCs may increase their signalling repertoire. We also believe that the appearance of NIT-GCs was one of the means to generate the diversity of signalling pathways using NO during the early stages of nervous system evolution in metazoans, since sponges and ctenophores, where NO signalling is present, have no NIT-GCs at all, as many as 12 have been found in placozoans, and NIT-GCs are functional in the "proto-eye" circuit of Platynereis.

>>>>

**Mathematical modelling of cPRC-circuit dynamics**

[to be added]

@@

The activity of synaptic circuits is richly modulated monoamines, neuropeptides and gaseous molecules. Our work revealed an essential role for NO-mediated signalling in driving UV/violet avoidance behaviour. NO, produced by postsynaptic interneurons, feeds back to their presynaptic photoreceptor partners leading to their delayed activation. This late-phase activation of cPRCs drives circuit output through projection interneurons and ciliomotor neurons. In the cPRC circuit, synaptic connectivity alone is not sufficient to drive circuit dynamics and behavioural change.

We identified cPRC-expressed NIT-GC1 as a new type of NO sensor. NIT-GC1 is localised to the axonal projections of the cPRCs in the anterior neurosecretory plexus of the brain. Here, cPRC axons intermingle and synapse on INNOS neurites that contain NOS. NO-mediated feedback signalling thus likely occurs in the neurosecretory plexus where NOS- and NIT-GC1-containing projections are in close proximity and where we could detect NO production following UV stimulation.

We showed that NIT-GC1 can produce cGMP in a NO-dependent manner. In cPRCs, increased cGMP levels could open CNGAα, a cyclic nucleotide-gated channels specifically expressed in cPRC [@Tosches2014], leading to cPRC delayed depolarisation, even after UV off.

NO is a free radical with a half-life of 5 msec in vivo and a signalling range limited to a few tens of nanometers. In neuronal signalling, NOS and sGC have been found to localise at close proximity at synapses [@burette2002][@garthwaite2015a]. We identified 12 sGCs in *Platynereis*, but none of these is expressed in the cPRC. Instead, we identified an unconventional NIT-domain containing GC, NIT-GC1 as the mediator of NO retrograde signalling. The NIT domain was first identified in bacteria and animal NIT-GCs have only been recently reported [@Shu2003; @Moroz2020]. Bacterial NIT domains regulate cellular functions in response to changes in extracellular and intracellular nitrate and/or nitrite concentrations [@camargo2007]. NO is readily converted to nitrate and nitrite [@santos2011; @möller2019; @garthwaite2015a] and these molecules accumulate in placozoans and cnidarians in cells and tissues with high NOS activity [@Moroz2004; @Moroz2020]. It is possible that the NIT domains in NIT-GCs may also sense nitrate and nitrite, as in bacteria. Furthermore, if they show different sensitivities to NO, nitrite and nitrate, it is possible that differences in their half-lives could give rise to a range of activation timings [@lundberg2011]. We have found 15 NIT-GCs in *Platynereis*, suggesting a wide range of functions. In addition, NIT-GC1 and NIT-GC2 showed very different subcellular localisation and functions. It is not clear whether the functional differences are due to the different localisation or also to differences in biochemical function. Overall, the diversification and spatially different localisation patterns of NIT-GCs may increase their signalling repertoire. Our data revealed key roles for two *Platynereis* NIT-GCs in phototraonsduction and NO-mediated neuromodulation suggesting that these molecules contribute to the diversity of neuronal signalling. Sponges and ctenophores, where NO signalling is present, lack NIT-GCs. In placozoans as many as 12 have been found, and cnidarians also contain a large diversity. NIT-GCs have been lost from chordates [check].

cPRCs expressing UV-absorbing c-opsin1 have high resting calcium levels and show hyperpolarising and subsequent depolarising responses to UV light (405 nm) [@Veraszto2018]. In the present study, we have shown that the cells that strongly synapse with cPRCs are interneurons that specifically express NOS (INNOS).

[@puopolo2001][@zhang2007][@Randel2014a]. In the mammalian retina, it is known that signals from bipolar photoreceptor cells are transmitted to downstream NOS-expressing amacrine cells (NOAC), where NO is produced [@jacoby2018]. These suggest that UV stimulation of cPRCs causes the production of NO via a depolarising response of INNOS.

In the mammalian retina, NO is an essential signalling molecule and one of the most well-studied [@Cudeiro1999]. It has been reported that NO produced in one of the retinal cells modulates cGMP-mediated neurotransmission by activating presynaptic neural sGCs (general target of NO) through retrograde signalling [@Vielma2014][@Wei2012].

Another type of interneuron that strongly synapses with the cPRC is the INRGWa, which synapses directly onto the ciliomotor neuron and is therefore considered to be a very important neuron in the control of larval behaviour [@Veraszto2018][@williams2017]. The present results suggest that the depolarising response of the cPRC due to NO feedback induces INRGW activation.

In addition, in many marine invertebrates, NO has been reported to be involved in many aspects of settlement behaviour and metamorphosis induction [@leise2001; @bishop2003; @bishop2007; @pechenik2007; @castellano2014; @ueda2013; @ueda2014; @ueda2016; @yang2018; @zhu2020; @song2021]. The downward swimming behaviour of Platynereis larvae is the first step in the onset of settlement behaviour and metamorphosis. These suggest that downward swimming due to NO activation of INRGWs may trigger the onset of settlement behaviour and metamorphosis [@Conzelmann2013][@Williams2015]. These suggest that NO feedback signalling may act as a trigger for downward swimming via activation of INRGWs, which induces the initiation of settlement behaviour and metamorphosis.
